---
title: "ED project"
author: "Jing Yeh, Saurav Sathnarayan"
date: "2024-09-03"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Contents

#### Introduction

#### Objectives

#### Sources of Variation

#### Randomisation

#### Analysis of pilot experiment

#### Our Model and Analysis

#### Summary and Conclusions

#### References

#### Apenddix

\

_things need to change and edit_
\
1. add page numbers
\
2. might need to add new pages/headings depending on project length.
\
3. Add title page?

\newpage



```{r include=FALSE}
#page 1
```

# Introduction

The goal of this experiment is to identify the programming language that delivers the fastest execution time when calculating a value of $\pi$ with respect to Leibiniz formula. $$\sum_{n=0}^{\infty} (-1)^n/(2n+1)$$

With the increasing demand for high-performance applications, understanding which programming languages offer superior speed in terms of execution is crucial for developers, especially in domains requiring real-time processing, large-scale data analysis, and resource-intensive computations.
\
This problem will focus on evaluating a selection of popular programming languages, including but not limited to C++, C, R, Python, Java, and JavaScript. The evaluation will consider how quickly a value of pi can be calculated using leibiniz formula, using varying number of terms, ie 100000 terms or 1000000 terms.We are not concerned about how accurate our value of $\pi$ is, but rather how quickly a programming language computes that value.
\

#### Compiled Language:
\
In a compiled language, the source code is translated into machine code by a compiler before execution. This machine code, often called an executable, can be run directly by the computer's hardware.
\
Compiled programs typically run faster since they are already in machine language, which the computer's processor can execute directly.
\
Examples: C, C++, Rust, and Go are examples of compiled languages.

#### Interpreted Language:
\
In an interpreted language, the source code is executed line-by-line by an interpreter at runtime. The interpreter reads the code, translates it into machine code, and executes it on the fly.
\
Interpreted programs generally run slower than compiled ones because the translation happens during execution.
\
Examples: Python, JavaScript, Ruby, and PHP are examples of interpreted languages.

#### Key Differences:
Compiled languages require a compilation step that produces an executable, while interpreted languages are executed directly by an interpreter.
\
 Compiled languages tend to have better performance due to the pre-compiled nature of the code, whereas interpreted languages are more flexible but slower due to the runtime translation.
\
Some languages, like Java, use a combination of both techniques, where the code is first compiled into an intermediate form (bytecode) and then interpreted or just-in-time (JIT) compiled at runtime.
\
_still need to edit this_


\newpage


```{r include=FALSE}
#page 2
```

# Objectives

_list our hypothesis here (is there a difference or not?)_
\
_list our contrasts. Figure out whether whihc method we are using to control type 1 error_
\
_what we hope to find from out experiment._

\newpage




```{r include=FALSE}
#page 3
```

# Sources of Variation

_discuss treatment factor and different levels_
\
_our experiment units, maybe list pc specs_
\
_go into depth about specs of pcs, blocking for each pc, to reduce experiment error variance._

\newpage




```{r include=FALSE}
#page 4
```

# Randomisation

_randomisation and soucres of variation can be one page_
\
_discuss process of randomisation, why we using it..._

\newpage


```{r include=FALSE}
#page 5
```

# Analysis of pilot experiment

_we need to decide on number of observations._
\
_full analysis here_
\
_list our hypothesis here (is there a difference or not?)_
\
_list our contrasts. Figure out whether which method we are using to control type 1 error_
\
_what we hope to find from out experiment._
\
_anova table and conclusions_
\
_contrasts and CI plus conclusions_

\newpage


```{r include=FALSE}
#page 6
```
# Our Model and Analysis

_list model formulation_
\
_explain model terms_
\
_repeat steps above_


\newpage


```{r include=FALSE}
#page 7
```
# Summary and Conclusions

_draw conclusions here_
\
_explain any challenges possibly_
\
_explain how we would do it differently, to fix errors or mistakes_
\newpage

```{r include=FALSE}
#page 8
```
# References
_any references, textbook, code, people wh helped us._
\newpage

```{r include=FALSE}
#page 9
```
# Apenddix
_add important code here._
