---
# This template is largely functionally equivalent to that for elsevier
# The two fields that are not in the elsevier format are:
#      `corresponding_author` and `acknowledgements`
title: STA2005S - Experimental Design Assignment
date: "`r Sys.Date()`"
author:
  - name: Jing Yeh
    email: yhxjin001@myuct.ac.za
    affiliation: University of Cape Town
  - name: Saurav Sathnarayan
    email: yhxjin001@myuct.ac.za
    affiliation: University of Cape Town

abstract: |
We aim to 
acknowledgements: |
  This is an acknowledgement.

  It consists of two paragraphs.
keywords:
  - key
  - dictionary
  - word
bibliography: mybibfile.bib
# Use a CSL when `citation_package = "default"`
#csl: https://www.zotero.org/styles/oxford-university-press-note

## Example of some pandoc's variable to use
#fontsize: 12pt
#spacing: halfline # could also be oneline
#classoptions:
#  - endnotes
#link-citations: yes
#urlcolor: orange
#linkcolor: green
#citecolor: red
#header-includes:
#  - \usepackage[nomarkers,tablesfirst]{endfloat} # For figures and tables at end
#  - \usepackage{lineno} # For line numbering
#  - \linenumbers # For line numbering

output: 
  rticles::oup_article:
    oup_version: 0 # 1 = 2020 CTAN OUP CLS package; 0 = 2009 OUP CLS package
    extra_dependencies: booktabs # for kable example
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE, # By default, hide code; set to TRUE to see code
  fig.pos = 'p', # Places figures on pages separate from text
  out.width = '100%', dpi = 300, # Figure resolution and size
  fig.env = "figure" # Latex figure environment
)
library(tidyverse)
library(scales)
library(gridExtra)
library(knitr)

```
\newpage

# Introduction
The goal of this experiment is to identify the programming language that delivers the fastest execution time when calculating a value of $\pi$ with respect to Leibiniz formula. $$\sum_{n=0}^{\infty} (-1)^n/(2n+1)$$

With the increasing demand for high-performance applications, understanding which programming languages offer superior speed in terms of execution is crucial for developers, especially in domains requiring real-time processing, large-scale data analysis, and resource-intensive computations.
\
This problem will focus on evaluating a selection of popular programming languages, including but not limited to C++, C, R, Python, Java, and Ruby. The evaluation will consider how quickly a value of pi can be calculated by applying leibiniz formula up to 100000000 terms.
\

#### Compiled Language:
\
In a compiled language, the source code is translated into machine code by a compiler before execution. This machine code, often called an executable, can be run directly by the computer's hardware.
\
Compiled programs typically run faster since they are already in machine language, which the computer's processor can execute directly.
\
Examples: C, C++, Rust, and Go are examples of compiled languages.

#### Interpreted Language:
\
In an interpreted language, the source code is executed line-by-line by an interpreter at runtime. The interpreter reads the code, translates it into machine code, and executes it on the fly.
\
Interpreted programs generally run slower than compiled ones because the translation happens during execution.
\
Examples: Python, JavaScript, Ruby, and PHP are examples of interpreted languages.

#### Key Differences:
Compiled languages require a compilation step that produces an executable, while interpreted languages are executed directly by an interpreter.
\
 Compiled languages tend to have better performance due to the pre-compiled nature of the code, whereas interpreted languages are more flexible but slower due to the runtime translation.
\
Some languages, like Java, use a combination of both techniques, where the code is first compiled into an intermediate form (bytecode) and then interpreted just-in-time (JIT) at runtime.
\
_still need to edit this_


\newpage

# Reference example

Here are two sample references:. Bibliography will appear at the end of the document.

# Methods
### Setting and Sample
This study was mostly conducted at the University of Cape Town, utilising the computers available on campus. We found that there are 5 different hardware setups available as experimental units. To supplement the range of our hardware setups, we also borrowed machines from our friends.

Since we did now the distribution of runtimes for various programming languages, we perform a priori tests to deduce that none of the languages has normally distributed runtime. To address this, we apply the Central Limit Theorem(CLT) to obtain a normal distribution for the average execution times. We ran the program 15 times per sample for each programming language, and repeated the process 30 times. Applying CLT, it is relatively safe to assume the distribution of sample means is approximately normal [2]. If we assume sample means to be normally distributed, the mean of the distribution of sample means is then an unbiased estimator for the true run time of each programming language[2], which we take as a single observation. This motivated us to design the experiment using Randomised Block Design (RBD)



### Design



An equation with a label for cross-referencing:

\begin{equation}\label{eq:eq1}
\sum_{n=0}^{\infty} (-1)^n/(2n+1)
\end{equation}

Since we don
```{r include FALSE}
#This equation can be referenced as follows: Eq. \ref{eq:eq1}
```
## A subsection

A numbered list:

1) First point
2) Second point
    - Subpoint
    
A bullet list:

* First point
* Second point

# Results
computers of different specifications are harder to come by, we will only use 3 different hardware setup for this pilot study.

## Pilot Study

### Problem: Runtimes of Programming Langugaes are not Normally Distributed
Upon inspecting the qqnorm plots of the runtime of all programming languages on the same machine, we discovered that the runtimes of languages clearly do not follow a normal distribution, even when the sample size is fairly large (n = 100)

To address this, we ran the program 15 times per sample for each programming language, and repeated the process 30 times. By the Central Limit Theorem (CLT), the distribution of sample means is approximately normal [2]. If we assume sample means to be normally distributed, the mean of the distribution of sample means is then an unbiased estimator for the true run time of each programming language[2], which we take as a single observation.

The process described above is automated using the following command:
```{eval=FALSE}
python3 run.py main 100000000 15 30
```


```{r, echo=FALSE, out.height="45%"}
setwd("./speed-comparison-master/data/")
ishango.before <- read.csv("ishango/noCLT/ishango.csv")
ishango.before.long <- pivot_longer(ishango.before, cols=everything(), names_to="language", values_to="runtime")
before <- ggplot(ishango.before.long, aes(sample=(runtime)))+
  stat_qq(col="green")+
  stat_qq_line(col="blue")+
  facet_wrap(~language, scales="free")+
  labs(x="quantiles", y="ms", title="Using One Sample of Size 100")+
  theme(axis.text.x = element_text(size=6))

ishango.after <- read.csv("ishango/fifteen30/ishango_mean.csv")

ishango.after.long <- pivot_longer(ishango.after, cols=everything(), names_to="language", values_to="runtime")
after <- ggplot(ishango.after.long, aes(sample=(runtime)))+
  stat_qq(col="coral")+
  stat_qq_line(col="blue")+
  facet_wrap(~language, scales="free")+
  labs(x="quantiles", y="ms", title="Using 30 Samples of Size 15")+
  theme(axis.text.x = element_text(size=6))

grid.arrange(before, after, ncol=2, nrow=1)
```


### Problem: Ensuring no Interactions between Blocking Factors and Treatment Factors
_todo: insert table for results_
```{r echo=FALSE}
setwd("./speed-comparison-master/data/")

pilot <- read.csv("pilotData.csv")
ttheme <- ttheme_minimal(base_size=10)
table <- tableGrob(pilot, them=ttheme)

pilot.long <- pivot_longer(pilot, cols=C:R, names_to = "language", values_to = "runtime")
plot <- ggplot(pilot.long, aes(x=language, y=log(runtime), group=Hardware, color=Hardware))+
  geom_point()+
  geom_line()+
  labs(title= "Interaction Graph",
       x= "Progamming Language",
       y= "Log Runtime")
grid.arrange(plot, table, ncol=1, nrow=2)
```
From the data collected, we observed that the results collected from Ishango do not follow the general trends established by the other three setups. Firstly, the hardware setup in Ishango lab is significantly less advance than MiddleTRNew. Yet, most programming languages tend to perform better on the Ishango machine. Secondly, to add to the first observation, not all programming languages perform better on the Ishango machine.

After further investigation, we learned that programming languages perform differently on various operating systems [4]. We hypothesised that this is likely the reason for the deviation, though further studies are needed to confirm this (we lack access to machines with the same hardware setup but run on different operating system).

Therefore, we added another constraint for selecting suitable machines: the machines must all run on Windows 10, as these machines are the most widely available.
## Generate a figure.

```{r fig1, fig.cap="This is the first figure.",echo=TRUE}
plot(1:10, main = "Some data", xlab = "Distance (cm)", ylab = "Time (hours)")
```

You can reference this figure as follows: Fig. \ref{fig:fig1}.

```{r fig2, fig.cap="This is the second figure.",echo=TRUE}
plot(1:5, pch = 19, main = "Some data", xlab = "Distance (cm)", ylab = "Time (hours)")
```

Reference to second figure: Fig. \ref{fig:fig2}

## Generate a table using `xtable`

```{r tab1, results="asis", echo=TRUE}
df <- data.frame(ID = 1:3, code = letters[1:3])

# Creates tables that follow OUP guidelines using xtable
library(xtable)
print(xtable(df, caption = "This is the table caption", label = "tab:tab1"),
  comment = FALSE
)
```

You can reference this table as follows: Table \ref{tab:tab1}.

## Generate a table using `kable`

```{r anothertable, echo=TRUE}
df <- data.frame(ID = 1:3, code = letters[1:3])

# kable can alse be used for creating tables
knitr::kable(df,
  caption = "This is the table caption", format = "latex",
  booktabs = TRUE, label = "tab2"
)
```

You can reference this table as follows: Table \ref{tab:tab2}.

# Discussion

You can cross-reference sections and subsections as follows: Section \ref{materials-and-methods} and Section \ref{a-subsection}.

***Note:*** the last section in the document will be used as the section title for the bibliography.

# References


# Appendix
#### PC Specificiations
\
\begin{tabular}{l|r|r|r}
\hline
  & CPU & MEMORY & OPERATING SYSTEM\\
\hline
Ishango PC &  9th Gen Intel® Core™ i3-9100  & 8,0 GB &  Ubuntu 22.04\\
\hline
MidddleTROld & 9th Gen Intel(R) Core(TM) i5-9500 CPU & 8,0 GB &  Windows 10\\
\hline
MiddleTRNew & 12th Gen Intel(R) Core(TM) i5-13400 & 16,0 GB &  Windows 10\\
\hline
ScilabB &  12th Gen Intel(R) Core(TM) i5-12400 &  16,0 GB  &  Windows 10\\
\hline
Surface &  9th Gen Intel(R) Core(TM) i5-8250 CPU &  16,0 GB  &  Windows 10\\
\hline
ASUS laptop &  5th Gen Intel(R) Core(TM) i7-5500U CPU &  6,0 GB  &  Windows 10\\
\end{tabular}

\

